Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-12-15 Manuscript/Manuscript.Rmd

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: Johanna Munoz Avila
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. 
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
output: rticles::jss_article
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

We talk of multilevel data when there is some kind of hierarchy or clustering in a dataset. In the typical case, individuals are nested within groups, but there are many different types of multilevel data. In the medical field, clustering occurs at e.g., the hospitals/center level in registry data, or at the study-level in meta-analyses (IPDMA). In the social sciences and official statistics we can find clustering e.g. at the country-level, or as imposed by the sampling design. In this paper, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit'.^[Alternatives to unit: participant/response/record.] For reasons of brevity, we only discuss clustering between units, not within units (such as in timeseries or longitudinal data).^[Also, add that we'll only discuss two levels, no more?]

Analyzing multilevel data requires special care, compared to 'regular', single level data. The cluster to which a unit belongs may influence the unit-level observations, and since clusters are made up of units, clusters depend on units as well [@hox17]. These relations can and should be taken into account when developing analysis models for multilevel data.^[Explain ICC here? The percentage of variance attributed to the cluster-level is expressed by the intra-class coefficient (ICC). The ICC can also be interpreted as the expected correlation between two randomly sampled units in same cluster. So if the ICC is high, a lot of variability in a variable is due to the clustering, which should be modeled accordingly.] Multilevel models typically include separate intercepts for each cluster, which relieves one restriction imposed by single-level models: equal group means across clusters. Additionally, there may be random predictor effects and/or random error terms (residual error variances), see e.g. @hox17 and @jong21.^[Add that heterogeneity refers to variability within clusters.] There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.

@@

```{r patterns, fig.height=4, fig.width=3.5, fig.cap = "Missingness in multilevel data", echo=FALSE}
plot_md_set()
# TODO: add border/line
# TODO: remove unit
# TODO: add accolade

@@

@@

Multilevel data is not spared the ubiquitous problem of missing information. Just as in single level data, missingness may occur at the unit level. But with multiple levels of data comes the potential for missingness at multiple levels. Missingness in multilevel data can be categorized into two general patterns: systematical missingness and sporadic missingness, see @resc13. In figure 1, we show a dataset with units in the rows and variables in the columns, there are 5 units nested within 2 clusters, and 3 variables of interest. Variable `X1` is completely observed. Variable `X2` is systematically missing, `X3` is sporadically missing.^[Explain why.] Systematic missingness can be further subdivided into unobserved constants (same value per cluster) and non-measured random variables (which may differ per unit within clusters). In Figure 1, the former implies that the unobserved values for units 4 and 5 on variable `X2` would be equal. With the latter, the values may differ. Depending on the missing data pattern, there are more or less optimal way of accommodating the missingness.

@@

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}. We provide practical guidelines and code snippets for different missing data situations. For reasons of brevity, we focus on imputation by chained equations, although JOMO is available in \pkg{mice} as well. Other useful packages for incomplete multilevel data include: \pkg{mitml}, \pkg{miceadds}, \pkg{mdmb}.^[Rephrase: Some level of knowledge on multilevel models is assumed. We're providing an overview of implementations. It's up-to the reader to decide which multilevel strategy suits their data. So we won't go into detail for the different methods (and equations). Refer to @meng94, @audi18, and @grun18. This paper is just a software tutorial. We'll keep it practical.] We introduce three case studies to illustrate the workflow.

@@

For each case study we focus on a different aspect to illustrate how to impute incomplete multilevel data. In the `mice::popmis` data, we show the advantages of including the multilevel structure of the data into the imputation model. In the `metamisc::impact` data we'll show how to induce missingness and solve it in real-world data. In the `GJRM::hiv` we provide novel methodology^[not really, the methods exist already, but how to show that this is something new and exciting?] for imputing MNAR missingness according to the Heckman model.

For each case study we'll look at least at: 1) the incomplete data; 2) the imputation model; 3) the imputed data; and 4) how to obtain pooled estimates for the analysis of scientific interest.

@@

# further inspection of the imputations
plot_box(imp_ignored, x = "popular", strip = TRUE)

>>>>

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: Johanna Munoz Avila
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  Multilevel data is not spared the ubiquitous problem of missing information. This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders.
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
output: rticles::jss_article
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

Research into any field with a hierarchical or clustered nature of observations may yield multilevel data. In the typical case, individuals are nested within groups, but there are many different types of multilevel data. In the medical field, clustering occurs at e.g., the hospital or center level in registry data, or at the study-level in meta-analyses (IPDMA^[remove or write in full?]). In the social sciences and official statistics we can find clustering e.g. at the country-level, or as imposed by a multi-stage sampling design. For the sake of legibility, we will refer to the grouping variable as 'cluster', and the grouped variable as '(sample) unit' throughout this paper.^[Alternatives to unit: participant/response/record.] And, for reasons of brevity, we only discuss clustering between units, not within units (e.g., timeseries or longitudinal data).^[Also, add that we'll only discuss two levels, not more?]

Analyzing multilevel data requires special care, compared to 'regular', single level data. For instance, there may be cross-level interactions between unit-level variables and cluster-level variables. The cluster to which a unit belongs may influence the unit-level observations, and vice versa for the the units that make up the cluster [@hox17]. These relations can and should be taken into account when developing analysis models for multilevel data.^[Explain ICC here? The percentage of variance attributed to the cluster-level is expressed by the intra-class coefficient (ICC). The ICC can also be interpreted as the expected correlation between two randomly sampled units in same cluster. So if the ICC is high, a lot of variability in a variable is due to the clustering, which should be modeled accordingly.] Multilevel models typically include separate intercepts for each cluster, which relieves one restriction imposed by single-level models: equal group means across clusters. Additionally, there may be random predictor effects and/or random error terms (residual error variances), see e.g. @hox17 and @jong21.^[Add that heterogeneity refers to variability within clusters.] There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.

@@

```{r patterns, fig.height=4, fig.width=3.2, fig.cap = "Missingness in multilevel data", echo=FALSE}
plot_md_set()
# TODO: add border/line
# TODO: remove unit
# TODO: add accolade

@@

Multilevel data is not spared the ubiquitous problem of missing information. Just as in single level data, missingness may occur at the unit level. But with multiple levels of data comes the potential for missingness at multiple levels. Missingness in multilevel data can be categorized into two general patterns: systematical missingness and sporadic missingness, see @resc13. In figure 1, we show a dataset with units in the rows and variables in the columns, there are 5 units nested within 3 clusters, and 3 variables of interest. Variable `X1` is completely observed. Variable `X2` is systematically missing, `X3` is sporadically missing.^[Explain why.] Systematic missingness can be further subdivided into unobserved constants (i.e., the same value within clusters) and non-measured random variables (which may differ per unit within clusters). In Figure 1, the former implies that the unobserved values for units 3 and 4 on variable `X2` would be equal. With the latter, the values may differ. Depending on the missing data pattern^[add missing data mechanisms here?], there are more or less optimal way of accommodating the missingness.

@@

@@

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}. We provide practical guidelines and code snippets for different missing data situations. For reasons of brevity, we focus on imputation by chained equations^[add that JOMO is available in \pkg{mice} as well?]. Other useful packages for incomplete multilevel data include \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}.^[Rephrase: Some level of knowledge on multilevel models is assumed. We're providing an overview of implementations. It's up-to the reader to decide which multilevel strategy suits their data. So we won't go into detail for the different methods (and equations). Refer to @meng94, @audi18, and @grun18. This paper is just a software tutorial. We'll keep it practical.]

@@

For each case study we focus on a different aspect to illustrate how to impute incomplete multilevel data. In the `mice::popmis` data, we show the advantages of including the multilevel structure of the data into the imputation model. In the `metamisc::impact` data we'll show how to induce missingness and solve it in real-world data. In the `GJRM::hiv` we provide novel methodology^[not really, the methods exist already, but how to show that this is something new and exciting?] for imputing MNAR missingness according to the Heckman model. For all case studies we discuss the nature of the incomplete data, the imputation model(s), and evaluation of the imputed data.

@@

# further inspection of the imputations
plot_box(imp_ignored, x = "popular", strip = TRUE)
# TODO: think about multimodality
